Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-26 submission/manuscript.Rmd
shigeki updates fig 2

Misexpression of germline genes in the adult brain suggests _Kdm5c_-KO cells fail to demarcate between germline and somatic identity. Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 3A, top). Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021]. Therefore, we assessed if KDM5C was necessary for embryonic silencing of germline genes by evaluating the impact of _Kdm5c_ loss in EpiLCs.

We first identified _Kdm5c_-KO EpiLC DEGs in of our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and assessed tissue-enriched gene expression. Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B). Using the curated list of germline genes generated above, we found _Kdm5c_-KO EpiLCs aberrantly expressed `r format(nrow(epigerm))` germline-enriched genes, including the previously characterized hippocampal DEG[@scandagliaLossKdm5cCauses2017] _Cytochrome C, testis-specific (Cyct)_ (Figure 3C). Although we observed aberrant expression of many tissue-enriched genes, we did not observe any significant difference in primed pluripotency genes (Figure 3D) or gross changes in _Kdm5c_-KO cell morpholgy during differentiation (Figure 3E), indicating KDM5C loss does not impair EpiLC formation.

We then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain to determine if all germline DEGs, like _Cyct_, are constitutively dysregulated or if they can change over the course of development. We found this was primarily not the case, as the majority of germline DEGs expressed in EpiLCs and the brain were unique, with only _Cyct_ shared across all sequencing datasets (Figure 3F). We then compared the known functions of EpiLC and brain germline DEGs and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle process (GO: 1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO: 0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes, such those involved in the sperm axoneme structure.

>>>>

Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 3A, top). While some germline-enriched genes are also expressed in embryonic stem cells (ESCs) and in the 2-cell stage[@wellingDAZLRegulatesTet12015; @macfarlanEmbryonicStemCell2012; @suzukiLossMAXResults2016], they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021]. Therefore, we tested if KDM5C was necessary for silencing germline genes in the post-implation embryo by evaluating the impact of _Kdm5c_ loss in EpiLCs.

We first identified _Kdm5c_-KO EpiLC DEGs in our previously published RNA-seq dataset[@samantaActivationXistEvolutionarily2022] (DESeq2, log2 fold change > `r l2fcco`, q < 0.1) and assessed tissue-enriched gene expression. Similar to the _Kdm5c_-KO brain, we observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3B). Although we observed aberrant expression of many tissue-enriched genes, including  `r format(nrow(epigerm))` germline-enriched genes, we did not find any significant difference in primed pluripotency genes, such as _Dnmt3b_, _Fgf5_, _Pou3f1_, and _Otx2_ (Figure 3C). We also did not observe any gross changes in _Kdm5c_-KO cell morpholgy during ESC to EpiLC differentiation (Figure 3D), altogether indicating KDM5C loss does not impair EpiLC formation.

To determine if germline DEGs are constitutively dysregulated or if they can change over the course of development, we compared compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain. We found the majority of germline DEGs expressed in EpiLCs and the brain were unique, with only _Cyct_ shared across all sequencing datasets (Figure 3E-F). We then evaluated the known functions of EpiLC and brain germline DEGs and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3G), such as meiotic cell cycle process (GO: 1903046, p.adjust = `r signif(getGO("meiotic cell cycle process", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`) and meiotic nuclear division (GO: 0140013, p.adjust = `r signif(getGO("meiotic nuclear division", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While there was modest enrichment of meiotic gene ontologies in in both brain regions, the _Kdm5c_-KO hippocampus primarily expressed late-stage spermatogenesis genes, such those involved in the sperm axoneme structure.
